Software Development in Startup Companies: The Green-field Startup Model

### 2013-02-19 article/main.tex
Changes before Carmines' last push

From a software engineering perspective development in startups is challenging since they develop software in a context where processes can hardly follow a prescriptive methodology \cite{Sutton2000, Coleman2005}. Despite startups share some characteristics with similar domains (e.g. small and web companies), the combination of different factors makes the specific development context quite unique \cite{Blank2005, Sutton2000}. Research is needed to support their engineering activities \cite{Coleman2005}, guiding practitioners in taking decisions and avoiding choices that could easily lead the whole business to failure \cite{Kajko-Mattsson2008}.
However, despite the impressive size of the startup ecosystem \cite{ISI:000243253000007}, the state-of-the-art presents a gap. Through a \textit{Systematic Mapping Study} (SMS) of the literature, conducted by Paternoster et al. \cite{SMS}, only few publications were found related to engineering activities in startups. Moreover the majority of these studies do not possess the relevance and rigour required to form a consistent body of knowledge in the state-of-the-art and in the state-of-practice.

@@

By means of the Green-field startup model (GSM), we captured the underlying phenomenon of software development in early-stage startups. The framework is based on a systematic procedure  which is grounded into empirical concepts obtained by conduction of the case-study. GSM presents the most significant themes of the development strategies that characterise startups' context.

The most urgent priority of software development is to shorten time-to-market. The greater part of engineering activities of startups are focused on the implementation while only little attention is given to more conventional activities (project management, requirement specifications, analysis, architecture design, automatic testing, \ldots).The first release of the product includes only a limited set of well suitable functionalities.

@@

The remain of this paper is structured as follow:  we provide relevant information about startups in section \ref{backg}. Section \ref{resmet} introduces the research questions and discusses the design and execution of the research process. Results are discussed in section \ref{res:gsm}, which presents the GSM model. Section \ref{res:val}  compares results of this research to literature and frameworks. Section \ref{valt} discusses the validity threats considered during this study. Finally, this paper concludes in section \ref{conc}.

@@

Modern entrepreneurship, which was born more than thirty years ago \cite{storey1982entrepreneurship}, has been boosted by the advent of the consumer internet markets in the middle of the nineties and culminated with the notorious \textit{dot-com bubble burst} of 2000 \cite{Perkins:1999:IBI:555126}. Several years later, with the massification of the internet and mobile devices, we are now assisting to an impressive proliferation of software ventures - metaphorically referred as the \textit{startup bubble}. The easy access to vast potential markets and the low cost of services distribution are appealing condition for modern entrepreneurs \cite{Marmer2011}. Inspired by stories of overwhelming successes, a large number of software businesses are created every day. However, the great majority of these companies fail within two years from their creation \cite{Crowne2002}.

@@

In this context, software engineering (SE) faces complex and multifaceted obstacles in understanding how to manage development processes. Sutton defines startups as creative and flexible in nature and reluctant to introduce process or bureaucratic measures which may result ineffective \cite{Sutton2000}. Also Bach refers to startups as \textit{``a bunch of energetic and committed people without defined development processes''} \cite{Bach1998}. In fact, startups have very limited resources and typically wish to use them to support product development instead of establishing processes \cite{Coleman2008, Heitlager2007}. Some attempts to taylor lightweight processes to startups reported basic failure of their application: \textit{``Everyone is busy, and software engineering practices are often one of the first places developers cut corners''} \cite{Martin2007}. Rejecting the notion of repeatable and controlled processes, startups prominently take advantage of unpredictable, reactive and low-precision \cite{surviving-os-cockburn} engineering practices \cite{Sutton2000, Tanabian2005, Chorev2006, Kakati2003}.

@@

The word \textit{startup} appeared in the SE literature for the first time in 1994 in an article written by Carmel \cite{Camel1994a} where he studied the \textit{time-to-completion in young package firm}. He noticed how these companies were particularly innovative and successful, advocating for the need of more research investigating software development practices so as to replicate success and try to transfer it to other technology sectors.

@@

A study of Kajko-Mattsson  \cite{Kajko-Mattsson2008}, which investigated a Swedish software startup, reported a heavy lack of requirements gathering process, minimal project management, lack of control over the change requests, absence of documentation to track the status and progress of the process and defective releases. Accordingly, Ambler et al. report how two startups approaching to an upcoming IPO started to require processes to focus on scalable solutions, in view of the growing company's size in terms of users and employees \cite{Ambler2002}. In this regard, Crowne, in \cite{Crowne2002}, specifies different stages through which software startups evolve. Starting without any established workflows, startups grow over time, creating and stabilizing processes to eventually improving them only when sufficiently mature.

But when startups have no time for training and orienting activities, as discussed in \cite{Sutton2000}, their main focus remains on team capabilities instead of prescriptive processes hiring people who can \textit{``hit the ground running''} \cite{Yoffie1999}. Empowering the team and focus on methodological attributes of the processes oriented in prototyping, proof-of-concepts, mock-ups and demos, to test basic functionalities, have been the primary priority in startups as described in \cite{Camel1994a}. Only when they grow, formal methodologies arise, followed by a conduction of quality assurance and long-term planning processes \cite{Yoffie1999}.

As studied in \cite{Tingling2007}, the maturity of a company affects the extent to which processes are adopted. The author reports how introducing Extreme Programming (XP) principles \cite{Beck:2004:EPE:1076267} in the development process was challenging because of the need of trained team-members for fully implementing the methodology. In fact, \cite{Silva2005} was able to start with all the XP practices in place only after six months of coaching the team, trying to enhance maturity from day-one. Nevertheless, even then, customization of practices were inevitably implemented to adapt the processes to the undertaken startups' context \cite{Deias}.

As partially discussed above, contributions to flexibility and reactiveness of the development process has been conducted prominently by means of \textit{Lean} \cite{Gautam2008} and \textit{Agile} \cite{Abrahamsson2002} methodologies (also reported in \cite{Taipale2010, Kuvinka2011}), where the extreme uncertain conditions lead startups to learn fast from trials and errors with a strong customer relationship in order to avoid wasting time in building wrong functionalities and prevent rapidly exhaustion of resources \cite{Midler2008, Hilmola2003, Sutton2000}. Customer involvement in software development has also been discussed in  \cite{Yogendra2002} as important factor to encourage an early alignment of business concerns to the technology strategies, because both are salient considerations to be successful \cite{Tingling2007}.

@@

When startups are mature enough to support software process improvement (SPI), the solutions considered according to the state-of-the-art are oriented towards light-weight processes such as a design of development process based on XP, proposed by Zettel in \cite{Zettel2001}. The process consists of a set of activities and artifacts (in addition to some important roles) defined in order to identify responsibilities and tools to utilize. But, despite the promising benefits reported by Zettel, we were not able to identify any future evaluation in real-world settings.

@@

The first publication mentioning the problem of \textit{one-size-fits-all}, related to the SPI  representations for startups, is described in \cite{Fayad1997}. The author reveals the problem in actuating the same \textit{best-practices} criteria for established companies in 10-person software startups. Thoroughly remarked in \cite{Sutton2000}, Sutton states that problems of SPI in software startups arise because of: the dynamism of the development process, which precludes repeatability; organizational maturity, that cannot be maintained from startups in view of lack of corporate direction; severe lack of resources, both human and technological for process definition, implementation, management, training \ldots; in conclusion, the primary benefits of SPI do not address startups, which instead of promoting product quality, aim to minimize time-to-market.

Additionally, the role of SPI has always been neglected because seen as an obstacle to the development team's creativity and flexibility as described in \cite{Coleman2008a} and to the need of a quick delivering product process environment. In fact, product quality is left aside in favour of minimal and suitable functionalities to shorten the time-to-market. As reported in two studies of Mater and Mirel \cite{Mater2000, Mirel2000}, quality aspects, mostly taken in consideration in internet startups, are oriented to usability and scalability, even though the market and application type heavily influences the quality-demand \cite{Coleman2008, Kim2005}.

Finally, to maintain the development activities, oriented to limited but suitable functionality, many studies suggest to externalize the complexity of parts of the project to third party solutions by means of outsourcing activities, software reuse and open-source strategies \cite{Hanna2010, Jansen2008, Wall2001, Bean2005}.

%--------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------%
\subsection{Technical debt}
Notwithstanding, a new interesting area of the SE research, trying to tackle the problem of the \textit{technical debt}, seems to bring interesting implications in studying development in software startups. The metaphoric neologism was originally introduced by Cunningham in 1992 \cite{TechnicalDebtCunn} and has recently attracted the attention of SE researchers\footnote{To attest the fact that \textit{technical debt} is gaining traction among researchers, we mention two important contributions which characterize the \textit{``debt landscape''} : \cite{Nugroho2011,Izurieta2012} and a dedicated workshop \cite{workshopDebt} organized by the Software Engineering Institute  and ICSE.}. The concept of \textit{technical debt} is well illustrated in \cite{Brown:2010:MTD:1882362.1882373} : \textit{``The idea is that developers sometimes accept compromises a system in one dimension (e.g., modularity) to meet an urgent demand in some other dimension (e.g., a deadline), and that such compromises incur a ``debt'' on which ``interest'' has to be paid and which the ``principal'' should be repaid at some point for the long-term health of the project''}.  The compromise between high-speed and high-quality engineering is faced daily by software startups, not only in terms of architecture design but in multifaceted aspects (weak project management, testing, process control, \ldots).

In conclusion, since \textit{``all decisions related to product development are trade-off situations''} \cite{Hilmola2003}, generally startups optimize workflows to the dynamic context they are involved into. In fact they typically adopt any development style that might work to support their first needs in what is called the \textit{``Just do it''} school of software startups \cite{Ries2011}. Additionally, as remarked by Coleman, \textit{``many managers just decide to apply what they know, as their experience tells them it is merely common sense''} \cite{Coleman2008}.

In this paper we make use of context-specific terminologies, which will be used to facilitate the readability and clarify any ambiguities throughout the whole paper in view of the weak state-of-the-art presented in \cite{SMS} :

@@

\begin{figure*}[!t]
\centering
\includegraphics[width=4in]{figures/completemethodology2}
\caption{Grounded theory study process overview}\label{fig:gt:completemethodology}
\end{figure*}

@@

Initially the boundaries of the research domain were set by means of non-systematic \textit{literature survey}, which provided the initial keywords used further in the research process. Moreover it helped us in defining the research questions (RQs):

@@

Despite GT has its roots in the social sciences, it has been increasingly used by \textit{Information System} researchers and has been almost ignored by the SE research \cite{Coleman2007}. In 1997 Bertelsen advocated for a need of more qualitative research in SE \cite{Bertelsen1997}, and in the last decade we assisted to an increasing number of publications that used a GT-like approach in SE \cite{Coleman2008a,Coleman2007, Sulayman2012}.

@@

\item The greatest part of engineering activities of startups are focused on the implementation while only little attention is given to more conventional activities (project management, requirement specifications, analysis, architecture design, automatic testing, \ldots). Especially when bringing the first product to the market, the attitude of startups towards traditional engineering activities is strongly characterized by a general feeling of unnecessity. Every document, artifact or process, which does not produce a short-term tangible benefit to the software production is perceived as a waste of time. The most visible downside of this kind of approach is a growing \textit{accumulated technical debt} (CAT6), which is partially mitigated by different elements supporting the development. Project management is substituted by the use of informal short-term deadlines supported by online collaborative tools with a kanban-board-like fashion; feature list are managed through ticket based systems supporting low-precision lists of requirement specifications in the form of user stories; after the essential set of functionality have been defined, making estimations and prioritization of features is not of much help; the traditional analysis process is, to a great extent, neglected (analysis is informally conducted for specific critical functionalities, usually by studying similar software solutions and rarely documenting decision-making processes); the design of the architecture is usually replaced by the use of well-known frameworks and naive diagrams on the whiteboard; advanced and modern version control systems work well for the configuration management; high-volumes of informal communication and the use of coding standards reduce the need of formal documentation; in the majority of startups writing automated test cases is perceived as a waste too (since the product reliability is not a primary concern,  engineers identify major bugs by testing the product internally and letting early adopters identify secondary bugs); finally, the infrastructural complexity of deployment is typically left to third party cloud applications.  Thus, by radically ignoring activities related to documentation, structures and processes, startups achieve a very high initial development speed in the first stage, but on the other hand they accumulate a technical debt that needs to be at least partially fulfilled when the company and the product grow.

@@

\item Engineering activities are supported by low-precision artifacts and collaborative tools integrated with the development environment. Startups take advantage of the simplicity of cutting-edge tools, which require a minimum learning and maintenance effort and provide a large payoff.  Tools are used for a vast number of tasks: ticked-based boards to manage user stories; systems to collect user feedback; tools for tracing product metrics; and scripts for managing deployments. The most prominent families of tools, which support software development in startups are modern version control systems. They are used by practitioners not only to manage the code base and the repositories, but for many other purposes such as: task assignment; tracing responsibilities; configuration management; issue management; automatic deployment; discussions and code reviews.  On the other hand, an excessive usage of  \textit{out-of-the-code} tools can negatively affect the development speed, as significantly reported by C11: \textit{``We realized we were losing too much time moving tickets around boards than actually working on stories. You need to find the balance between the time spent with tools and the amount of actual coding''}. This risk is often mitigated with an intense usage of offline manual tools: especially post-it notes and whiteboards are reported to be particularly effective to foster internal communication within the team.

@@

Finally, also the experience of team-members is considered as an important confounding factor. Since only little evidence has been gathered during the interviews, ``lack of experience'' was not considered as a sub-category in \textit{severe lack of resources}. Notwithstanding, according to \cite{Crowne2002}, startups at the beginning rely on clever, but inexperienced developers. In fact, having team-members with deep experience is a ``double-edge sword'', as discussed in \cite{Yoffie1999}. From one hand experience might quickly provide structure and maturity to the development process, from the other it might cause huge challenges in managing human resources in view of the fact that self-confident overachievers will almost inevitably clash. Consequently the team management requires higher control and coordination activities that inevitably hinder the flexibility of the development environment at the beginning, that is essential in early-stage startups. Accordingly, also Coleman in \cite{Coleman2007,Coleman2008a,Coleman2008} states that the operating strategies towards \textit{``minimum processes''} in startups is not a matter of poor knowledge and training, but rather it is the necessity of operating with solutions that let the company move faster.

>>>>

From a software engineering perspective development in startups is challenging since they develop software in a context where processes can hardly follow a prescriptive methodology \cite{Sutton2000, Coleman2005}. Despite startups share some characteristics with similar domains (e.g. small and web companies), the combination of different factors makes the specific development context quite unique \cite{Blank2005, Sutton2000}. Therefore, research is needed to support their engineering activities \cite{Coleman2005}, guiding practitioners in taking decisions and avoiding choices that could easily lead the whole business to failure \cite{Kajko-Mattsson2008}.
However, despite the impressive size of the startup ecosystem \cite{ISI:000243253000007}, the state-of-the-art presents a gap. Through a \textit{Systematic Mapping Study} (SMS) of the literature, conducted by Paternoster et al. \cite{SMS}, only few publications were found related to engineering activities in startups. Moreover the majority of these studies do not possess the relevance and rigour required to form a consistent body of knowledge in the state-of-the-art and in the state-of-practice.

@@

By means of the Green-field startup model (GSM), we captured the underlying phenomenon of software development in early-stage startups. The framework is based on a systematic procedure  which is grounded into empirical concepts obtained by a case-study. GSM presents the most significant themes of the development strategies that characterise startups' context.

The results of our study suggest that the most urgent priority of software development in startups is to shorten time-to-market. The greater part of engineering activities are focused on the implementation while only little attention is given to more conventional activities (project management, requirement specifications, analysis, architecture design, automatic testing).The first release of the product includes only a limited set of well suitable functionalities.

@@

The remainder of this paper is structured as follows:  we provide relevant information about startups in section \ref{backg}. Section \ref{resmet} introduces the research questions and discusses the design and execution of the research process. Results are discussed in section \ref{res:gsm}, which presents the GSM model. Section \ref{res:val}  compares results of this research to literature and frameworks presented in this field. Section \ref{valt} discusses the validity threats considered during this study. Finally, this paper concludes in section \ref{conc}.

%MUN: This is maybe only a minor concern, but it may be relevant for certain reviewers: The references you use
% in the introduction are not very recent. Look at the list of papers [1-15], most of them are published before 2005.
% I understand that you use some references to strengthen your general arguments (which can refer to older publications),
% but your introduction should also show the relevance of your research right now. This is best shown by referring to
% recent (max 5 years old) literature. Think about if you can add more recent (2010-now), well published, papers in the
% introduction.

@@

Modern entrepreneurship, which was born more than thirty years ago \cite{storey1982entrepreneurship}, has been boosted by the advent of the consumer internet markets in the middle of the 1990s and culminated with the notorious \textit{dot-com bubble burst} of 2000 \cite{Perkins:1999:IBI:555126}. Several years later, with the massification of the internet and mobile devices, we are now assisting to an impressive proliferation of software ventures - metaphorically referred as the \textit{startup bubble}. The easy access to vast potential markets and the low cost of services distribution are appealing condition for modern entrepreneurs \cite{Marmer2011}. Inspired by success stories, a large number of software businesses are created every day. However, the great majority of these companies fail within two years from their creation \cite{Crowne2002}.

@@

In this context, software engineering (SE) faces complex and multifaceted obstacles in understanding how to manage development processes. Sutton defines startups as creative and flexible in nature and reluctant to introduce process or bureaucratic measures which may result ineffective \cite{Sutton2000}. Also Bach refers to startups as \textit{``a bunch of energetic and committed people without defined development processes''} \cite{Bach1998}. In fact, startups have very limited resources and typically wish to use them to support product development instead of establishing processes \cite{Coleman2008, Heitlager2007}. Some attempts to taylor lightweight processes to startups reported basic failure of their application: \textit{``Everyone is busy, and software engineering practices are often one of the first places developers cut corners''} \cite{Martin2007}. Rejecting the notion of repeatable and controlled processes, startups prominently take advantage of unpredictable, reactive and low-precision \cite{
surviving-os-cockburn} engineering practices \cite{Sutton2000, Tanabian2005, Chorev2006, Kakati2003}.

@@

The word \textit{startup} appeared in the SE literature for the first time in 1994 in an article written by Carmel \cite{Camel1994a} where he studied the \textit{time-to-completion in a young package firm}. %MUN: If this is a quote from
%Carmel, put it in quotation marks, like you did in the previous section with Suttons quote. Generally, check that you
%are consistent how you use quotations throughout the paper.
He noticed how these companies were particularly innovative and successful, advocating for the need of more research investigating software development practices so as to replicate success and try to transfer it to other technology sectors.

@@

A study of Kajko-Mattsson \cite{Kajko-Mattsson2008}, which investigated a Swedish software startup, reported a heavy lack of requirements gathering processes, minimal project management, lack of control over the change requests, absence of documentation to track the status and progress of the process, and defective releases. Accordingly, Ambler et al. report how two startups approaching to an upcoming IPO started to require processes to focus on scalable solutions, in view of the growing company's size in terms of users and employees \cite{Ambler2002}. In this regard, Crowne \cite{Crowne2002} specifies different stages through which software startups evolve. Starting without any established workflows, startups grow over time, creating and stabilizing processes to eventually improving them only when sufficiently mature.

But when startups have no time for training and orienting activities, as discussed in \cite{Sutton2000}, their main focus remains on team capabilities instead of prescriptive processes, hiring people who can \textit{``hit the ground running''} \cite{Yoffie1999}. Empowering the team and focus on methodological attributes of the processes oriented in prototyping, proof-of-concepts, mock-ups and demos, to test basic functionalities, have been the primary priority in startups \cite{Camel1994a}. Only when they grow, formal methodologies arise, followed by performing quality assurance and long-term planning processes \cite{Yoffie1999}.

The maturity of a company affects the extent to which processes are adopted. Tingling \cite{Tingling2007} reports how introducing Extreme Programming (XP) principles \cite{Beck:2004:EPE:1076267} in the development process was challenging because of the need of trained team-members for fully implementing the methodology. Da Silva and Kon \cite{Silva2005} were able to start with all the XP practices in place only after six months of coaching the team, trying to enhance maturity from day-one. Nevertheless, even then, customization of practices were inevitably implemented to adapt the processes to the undertaken startups' context \cite{Deias}.

As partially discussed above, contributions to flexibility and reactiveness of the development process has been conducted prominently by means of \textit{Lean} \cite{Gautam2008} and \textit{Agile} \cite{Abrahamsson2002} methodologies (also reported in \cite{Taipale2010, Kuvinka2011}), where the extreme uncertain conditions lead startups to learn fast from trials and errors with a strong customer relationship in order to avoid wasting time in building wrong functionalities and prevent rapidly exhaustion of resources \cite{Midler2008, Hilmola2003, Sutton2000}. Customer involvement in software development has also been discussed by Yogendra \cite{Yogendra2002} as an important factor to encourage an early alignment of business concerns to the technology strategies, because both are salient considerations to be successful \cite{Tingling2007}.

@@

When startups are mature enough to support software process improvement (SPI), the solutions considered according to the state-of-the-art are oriented towards light-weight processes such as a design of development process based on XP, proposed by Zettel \cite{Zettel2001}. The process consists of a set of activities and artifacts (in addition to some important roles) defined in order to identify responsibilities and tools to utilize. But, despite the promising benefits reported by Zettel, we were not able to identify any future evaluation in real-world settings.

@@

The problem of \textit{one-size-fits-all}, related to the SPI  representations for startups, is described by Fayad \cite{Fayad1997}. The author discusses the problem in actuating the same \textit{best-practices} criteria for established companies in 10-person software startups. Thoroughly remarked in \cite{Sutton2000}, Sutton states that problems of SPI in software startups arise because of: the dynamism of the development process, which precludes repeatability; organizational maturity, that cannot be maintained from startups in view of lack of corporate direction; severe lack of resources, both human and technological for process definition, implementation, management, and training. In conclusion, the primary benefits of SPI do not address startups, which instead of promoting product quality, aim to minimize time-to-market.

Additionally, the role of SPI has been neglected because seen as an obstacle to the development team's creativity and flexibility \cite{Coleman2008a}, and to the need of a quick delivering product process environment. %MUN: the second
%part of the sentence is just a statement without reference. If it refers to a statement by Coleman, put his reference
%at the end OR find the correct reference OR delete the statement.
Product quality is left aside in favor of minimal and suitable functionalities to shorten the time-to-market. As reported in two studies of Mater and Mirel \cite{Mater2000, Mirel2000}, quality aspects, mostly taken in consideration in internet startups, are oriented to usability and scalability, even though the market and application type heavily influences the quality-demand \cite{Coleman2008, Kim2005}.

To maintain the development activities, oriented to limited but suitable functionality, studies suggest to externalize the complexity of parts of the project to third party solutions by means of outsourcing activities, software reuse and open-source strategies \cite{Hanna2010, Jansen2008, Wall2001, Bean2005}. %MUN: Can you split up the references? I.e. write:
%"outsourcing activities \cite{Bla}, software reuse \cite{Bla}, and open-source strategies \cite{Bla}.

%--------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------%
\subsection{Technical debt}
Notwithstanding, a new interesting area of the SE research, trying to tackle the problem of the \textit{technical debt}, seems to bring interesting implications in studying development in software startups. The metaphoric neologism was originally introduced by Cunningham in 1992 \cite{TechnicalDebtCunn} and has recently attracted the attention of SE researchers\footnote{To attest the fact that \textit{technical debt} is gaining traction among researchers, we mention two important contributions which characterize the \textit{``debt landscape''} : \cite{Nugroho2011,Izurieta2012} and a dedicated workshop \cite{workshopDebt} organized by the Software Engineering Institute  and ICSE.}. The concept of \textit{technical debt} is well illustrated in \cite{Brown:2010:MTD:1882362.1882373} : \textit{``The idea is that developers sometimes accept compromises in a system in one dimension (e.g., modularity) to meet an urgent demand in some other dimension (e.g., a deadline), and that such compromises incur a ``debt'' on 
which ``interest'' has to be paid and which the ``principal'' should be repaid at some point for the long-term health of the project''}.  The compromise between high-speed and high-quality engineering is faced daily by software startups, not only in terms of architecture design but in multifaceted aspects (weak project management, testing, process control).

Since \textit{``all decisions related to product development are trade-off situations''} \cite{Hilmola2003}, generally startups optimize workflows to the dynamic context they are involved into. They typically adopt any development style that might work to support their first needs in what is called the \textit{``Just do it''} school of software startups \cite{Ries2011}. Additionally, as remarked by Coleman, \textit{``many managers just decide to apply what they know, as their experience tells them it is merely common sense''} \cite{Coleman2008}.

\subsection{Terminology}
In order to set a common ground and to prevent ambiguity, we use the following terminology throughout the paper:

@@

\begin{figure*}[!t]
\centering
\includegraphics[width=4in]{figures/completemethodology2}
\caption{Grounded theory study process overview}\label{fig:gt:completemethodology}
\end{figure*}
%MUN: I think you can increase the font size in the figure a bit. Try thereby to keep the 
%height of the figure, expanding however the width (use horizontal space since it is a two 
%column figure).
%Write "Data analysis" horizontally

@@

Initially the boundaries of the research domain were set by means of non-systematic literature survey, which provided the initial keywords used further in the research process. Moreover it helped us in defining the research questions (RQs):

@@

Despite GT has its roots in the social sciences, it has been increasingly used by Information System researchers and has been almost ignored by the SE research \cite{Coleman2007}. In 1997 Bertelsen advocated for a need of more qualitative research in SE \cite{Bertelsen1997}, and in the last decade we assisted to an increasing number of publications that used a GT-like approach in SE \cite{Coleman2008a,Coleman2007, Sulayman2012}.

@@

\item The greatest part of engineering activities of startups are focused on the implementation while only little attention is given to more conventional activities (project management, requirement specifications, analysis, architecture design, automatic testing, \ldots). Especially when bringing the first product to the market, the attitude of startups towards traditional engineering activities is strongly characterized by a general feeling of unnecessity. Every document, artifact or process, which does not produce a short-term tangible benefit to the software production is perceived as a waste of time. The most visible downside of this kind of approach is a growing \textit{accumulated technical debt} (CAT6), which is partially mitigated by different elements supporting the development. Project management is substituted by the use of informal short-term deadlines supported by online collaborative tools with a kanban-board-like fashion; feature list are managed through ticket based systems supporting low-
precision lists of requirement specifications in the form of user stories; after the essential set of functionality have been defined, making estimations and prioritization of features is not of much help; the traditional analysis process is, to a great extent, neglected (analysis is informally conducted for specific critical functionalities, usually by studying similar software solutions and rarely documenting decision-making processes); the design of the architecture is usually replaced by the use of well-known frameworks and naive diagrams on the whiteboard; advanced and modern version control systems work well for the configuration management; high-volumes of informal communication and the use of coding standards reduce the need of formal documentation; in the majority of startups writing automated test cases is perceived as a waste too (since the product reliability is not a primary concern,  engineers identify major bugs by testing the product internally and letting early adopters identify secondary 
bugs); finally, the infrastructural complexity of deployment is typically left to third party cloud applications.  Thus, by radically ignoring activities related to documentation, structures and processes, startups achieve a very high initial development speed in the first stage, but on the other hand they accumulate a technical debt that needs to be at least partially fulfilled when the company and the product grow.

@@

\item Engineering activities are supported by low-precision artifacts and collaborative tools integrated with the development environment. Startups take advantage of the simplicity of cutting-edge tools, which require a minimum learning and maintenance effort and provide a large payoff.  Tools are used for a vast number of tasks: ticked-based boards to manage user stories; systems to collect user feedback; tools for tracing product metrics; and scripts for managing deployments. The most prominent families of tools, which support software development in startups are modern version control systems. They are used by practitioners not only to manage the code base and the repositories, but for many other purposes such as: task assignment; tracing responsibilities; configuration management; issue management; automatic deployment; discussions and code reviews.  On the other hand, an excessive usage of  \textit{out-of-the-code} tools can negatively affect the development speed, as significantly reported by C11: \
textit{``We realized we were losing too much time moving tickets around boards than actually working on stories. You need to find the balance between the time spent with tools and the amount of actual coding''}. This risk is often mitigated with an intense usage of offline manual tools: especially post-it notes and whiteboards are reported to be particularly effective to foster internal communication within the team.

@@

Finally, also the experience of team-members is considered as an important confounding factor. Since only little evidence has been gathered during the interviews, ``lack of experience'' was not considered as a sub-category in \textit{severe lack of resources}. Notwithstanding, according to \cite{Crowne2002}, startups at the beginning rely on clever, but inexperienced developers. In fact, having team-members with deep experience is a ``double-edge sword'', as discussed in \cite{Yoffie1999}. From one hand experience might quickly provide structure and maturity to the development process, from the other it might cause huge challenges in managing human resources in view of the fact that self-confident overachievers will almost inevitably clash. Consequently the team management requires higher control and coordination activities that inevitably hinder the flexibility of the development environment at the beginning, that is essential in early-stage startups. Accordingly, also Coleman in \cite{Coleman2007,
Coleman2008a,Coleman2008} states that the operating strategies towards \textit{``minimum processes''} in startups is not a matter of poor knowledge and training, but rather it is the necessity of operating with solutions that let the company move faster.
